---
title: "Testing the CNN results"
author: "Adela Sobotkova"
date: "November 24, 2021 updated `r format(Sys.time(), '%B %d, %Y')`" 
output: 
  html_document:
    toc: true
    toc_float: true
    highlight: tango
    code_folding: hide
---

## Libraries
```{r libraries}
library(tidyverse)
library(sf)
library(raster)
```

## Load Data

```{r}
objects_needed <- c("kaz", "cnn_df", "grid60", "mounds", "survey_ch", "survey", "survey_grid60")
script <- c("CNN_LoadData2021.R")

for (i in objects_needed){
  if (exists(i)){
      print(i)
  }  else {
    source(script)
  #  source(paste0("scripts/",(script[contains(gsub("m","", i), vars = script)])))
  }
}
```

## Plot the data
```{r view}
# View all mounds
plotRGB(kaz, stretch= "lin");
plot(survey_ch, border = "green", lwd = 2, add = TRUE);
#plot(survey_sm, border = "red", lwd = 2, add = TRUE);
#plot(survey$geometry, col = "lightyellow",border = "lightyellow", add = TRUE );
plot(mounds$geometry, add = TRUE, col = "hotpink");
plot(grid80$geometry, add = TRUE, col = "white");
plot(grid60$geometry, add = TRUE, border = "white");
legend("bottom", legend = "Grid cells with 60% (white border) and 80% (white fill) probability of mounds overlayed on \nthe satellite image and survey study area outlines with mounds (pink circles)")

```
## Validation

The Tundzha Regional Archaeological project surveyed 85 sq km of the area included in the satellite imagery in 2009-2011, documenting the burial mounds within which create a dataset suitable for validation.

Area surveyed contains 773 documented mounds of various shapes and sizes, circa half of them visible on surface.
The written comments for chunks below are written for the original mound_probability counts in CNN01_LoadData2021.R.

Let's explore how many of these mounds are inside and outside grid cells with 60 and 80% likelihood. Bring in mound attributes and check correlation. Does size play a role?



## True Positives at 60% threshold
Archaeological survey documented 773 mounds in the study area. 37 mounds from this dataset that fall in the 60%+ probability gridcells form the true positives and comprise the success of the CNN model.
```{r overlap}
# Check spatial overlap between 60%+ grid cells and all 773 mounds 
predicted60 <- st_intersection(mounds, grid60)# %>% distinct() # 37 predicted

# Which ones are not-predicted? 
`%nin%` = Negate(`%in%`)
unpredicted60 <- mounds[mounds$TRAP_Code%nin%predicted60$TRAP_Code,] # 736 not predicted

# Inspect
length(unique(predicted60$TRAP_Code)) # 37 unique mounds
length(grid60$id) # 270 cells with 60%p in the whole Kazanlak IKONOS
length(survey_grid60) #134 cells with 60%p in survey area

# How many gridcells with 60% probability actually contain mounds? 
grids_n_mounds <- predicted60 %>% 
 #st_drop_geometry() %>% 
  group_by(id) %>%
  count() 
mapview(grids_n_mounds) 

length(unique(grids_n_mounds$id)) # 50 grids within study area catch mounds

hist(grids_n_mounds$n, breaks = 40, 
     xlab = "Mound count in grid",
     ylab = "Grid count",
     main = paste0("Number of predicted mounds (n = ",length(unique(predicted60$TRAP_Code)), ") per grid cell ( ",nrow(grids_n_mounds)," TP) \n among a total of ",length(survey_grid60)," grid cells in the study area with 60% likelihood"))

# Summary: 50 out of 332 Grid cells with 60%+ probability contain between 1 and 29 detected mounds, which makes sense for the necropolis of little 10m diameter mounds in the NW
```

The previous chunk shows that 37 of 773 unique mounds appear in 18 of the 134 deduplicated "survey" (`r nrow(grids_n_mounds)/length(survey_grid60)`gridcells are TP) grids that have 60% or higher probability of mounds and are at least partially within the TRAP study area. 

In summary:
The entire satellite image contains `length(grid60$id)` 270 grids with 60%+ probability of mounds, 134 fall in the TRAP study area and 18 of these contain 37 mounds.


### Write prediction status to mounds
```{r write-predicted60-status}
# Add the information on prediction to mound dataset
library(tidyverse)
mounds <- mounds %>% 
  mutate(predicted60 =  case_when(TRAP_Code %in% predicted60$TRAP_Code ~ "yes",
                                  TRAP_Code %nin% predicted60$TRAP_Code ~ "no"
                                  ))
```


## Visualize results 
```{r mapview}
# See the predicted mounds over grids with 60% likelihood of mound plus unpredicted mounds

library(mapview)
mapview(unpredicted60, color = "orange", alpha= 0.5,
         map.types = c("Esri.WorldImagery", "CartoDB.Positron"),
         layer.name = "Undetected mounds")+
  mapview(grids_n_mounds, zcol = "n", at = c(1, 2, 5, 10, 15, 30),
          layer.name = "Detected mound no per gridcell")+
  mapview(survey_grid60, layer.name = "Gridcells with 60%+ probability")
```

## False Negatives at 60% threshold
Given 38 mounds were predicted by CNN, where are the remaining 735?
Let's explore where the missing mounds are and what probability has been assigned to the cells that contain them by CNN model.

```{r find-missing-I-problematic}
# Find missing mound information regarding prediction in the overall grid cnn_all
# Need to eliminate overlap in gridcells!
# eliminating id duplicates is nonsensical as the two datasets have same ids, I  must aggregate spatially, e.g. st_union or st_difference, or relabel the datasets before merging, or crop to a shared non-overlapping border.
missing60 <- st_intersection(unpredicted60, cnnall_grid) # 2158, clearly duplicates

# Check for duplicates
#?distinct()
length(duplicated(missing60$TRAP_Code)) # 2158 points to duplication
missing60 <- missing60[!duplicated(missing60$TRAP_Code),] #745 undetected mounds
```

```{r find-missing-I}
# Find missing mound information regarding prediction in the overall grid cnn_all
# Need to eliminate overlap in gridcells!
# eliminating id duplicates is nonsensical as the two datasets have same ids, I  must aggregate spatially, e.g. st_union or st_difference, or relabel the datasets before merging, or crop to a shared non-overlapping border.
missing60 <- mounds %>%
             dplyr::filter(TRAP_Code %nin% predicted60$TRAP_Code) # 752, clearly duplicates

unpredicted60 # 736

# Check for duplicates
sum(duplicated(unpredicted60$TRAP_Code)) # no duplicates

# Tile success rates
all_tiles <- st_intersection(mounds, cnnall_grid) %>% group_by(id) %>% count()
fp_tiles <- st_intersection(unpredicted60, cnnall_grid)
fp_tiles %>% group_by(id) %>% count()

```

```{r missing-in-grids}
# How many grids are the missing mounds in?
missing60 %>% 
  st_drop_geometry() %>% 
  group_by(X1) %>% 
  tally() %>% arrange(desc(n))
# The missing 620 mounds are contained within 259 grid cells of the satellite image 

# Let's see the distribution of missing mounds per grid cell
missing60 %>% 
  group_by(X1) %>% 
  count() %>% 
  ggplot()+ geom_histogram(aes(n))+ggtitle("Count of undetected mounds (620) per gridcell(n=259)") + theme_bw()
# some gridcells contain up to 28 undetected mounds. No surprise with the NW necropolis
```

Undetected mounds appear in 259 gridcells, which contain between 1 and 29 mounds, heavily skewed towards small mound number within cell. 

What is the probability of the cells that contain undetected mounds?
```{r missing-prob}
# What is the mound_probability in these gridcells?
hist(missing60$mound_probability, main = "Probability of gridcells containing undetected mounds")
```


### Explore impact of height on detectability
Let's look at whether mound Height is a factor impacting detectability. (it is a factor for visual inspection as higher mounds are more visible in sat img.) 
```{r explore-height}
# filter the mounds for largesh and noticeable ones (2+ meters)
large <- mounds %>% 
  filter(Height>=2) # 247 obs

mounds %>% 
  group_by(predicted60) %>% 
  summarize(min = min(Height), max = max(Height))

# Check the height difference between un- and predicted mounds
boxplot(Height~predicted60, data = mounds, xlab = "Was this field-confirmed mound detected with CNN?")

# Check height distribution btw un- and predicted mounds
hist(mounds$Height[mounds$predicted60 == "no"], col = "pink", breaks  = 20,main = "Histogram of detected (yellow) and undetected (pink) mounds by height",
     xlab = "Mound height (m)");
hist(mounds$Height[mounds$predicted60 == "yes"], col = "yellow", add =TRUE, alpha = 0.5);
```

### T-test on height
```{r ttest-height}
# Is there a significant relationship between the Height of a mound and its detection?
t.test(Height ~ as.factor(predicted60), data = mounds)
str(t.test(Height ~ as.factor(predicted60), data = mounds))
# ... the answer is inconclusive with p at 0.483
```

## Summary of success rates

38 mounds was detected out of 773, making the CNN predict ca 5% of mounds documented in the study area. Out of the 151 gridcells which are within the archaeological study area and have 60%+ probability of containing a mound, 19 (12%) actually contained between 1 and 10 mounds. 

The remaining undetected 735 mounds were found in XXX gridcells, clustered from 1 to 28 specimen. 
The explored numbers (both true and FP) may be further inflated as there is an underlying grid overlap which cause reduplication of probability values (which are not identical) and proliferation of mound duplicates which each have different probability value! See section `discrepancy` above.


### What is the model succeeding at and what not?
Model IS detecting the reservoir and mostly avoiding it. Model is picking up the NE cluster (correctly) and SW cluster (less correctly).
In NE we are missing a lot of both small and very large 20m diameter+ mounds that are covered with scrub (look forested) . Strangely the algorithm picked one place without mounds - which is a feat, given the high mound density in this area.
In the south a lot of forest patches and beach boundaries are getting picked up that are false positives. Here the overlap may be at play !
All in all the model seems to pick on the bright signatures rather than round shapes and there is an odd combination of large forested lozenges and small bare spots picked up, but somehow round forested images with little excavated (bright) trenches on the south side is not getting picked up.



## Additional visualisations

```{r viz-mounds-cnn}

# Where are the mounds vis-a-vis the probability values
cnnall_grid %>% 
  ggplot()+
  geom_sf(aes(color = mound_probability))+
  scale_color_gradient(low = "green", 
                       high = "red",
                       "Mound probability")+
  geom_sf(data = mounds) +
  ggtitle("Location of known mounds vis-a-vis the CNN predictions")


# See the mound features predicted with 60%+ probability 
grid60 %>% 
 # filter(mound_probability>0.5) %>% 
  ggplot()+
  geom_sf(aes(color = mound_probability))+
  scale_color_gradient(low = "green", 
                       high = "red",
                       "Mound probability")+
  geom_sf(data = mounds$geometry[mounds$predicted60 == "yes"], size = mounds$Height[mounds$predicted60 == "yes"], col = "lightgrey", alpha = 0.5) +
  ggtitle("Predicted mounds in Kazanlak")

# See the mounds not caught by the cells with 60%+ probability 
grid60 %>% 
  # filter(mound_probability>0.5) %>% 
  ggplot()+
  geom_sf(aes(color = mound_probability))+
  scale_color_gradient(low = "green", 
                       high = "red",
                       "Mound probability")+
  #geom_sf(data = unpredicted60$geometry)+
  geom_sf(data = mounds$geometry[mounds$predicted60 == "no"], size = mounds$Height[mounds$predicted60 == "no"], col = "lightgrey", alpha = 0.5)+
  ggtitle("Unpredicted mounds in Kazanlak")
```

## Side-by-side views 
These images take mound height into account and split the results into facets
```{r ggplot-facet}
# Cumulative facetted visual

ggplot()+
  geom_sf(data = grid60, aes(fill = mound_probability)) +
  labs(fill = "Probability of mound")+
  # add mounds
  geom_sf(data = mounds, 
          aes(size = Height, col = "red",  alpha = 0.6)) +
  facet_wrap(~predicted60)+
  theme_bw() + 
  guides(color = FALSE,
         size = guide_legend(order = 1),
         fill = guide_legend(order = 2),
         alpha = FALSE)+
   ggtitle(paste0("Mounds located in gridcells with 60%+ 
          probability (no=",nrow(unpredicted60),",yes=",nrow(predicted60),")"))
```

##True Positives at 80% threshold 

```{r overlap-80}
# Check spatial overlap between 80%+ grid cells and all 773 mounds 
predicted80 <- st_intersection(mounds, grid80)# %>% distinct() # 5 predicted

# Which ones are not-predicted? 
`%nin%` = Negate(`%in%`)
unpredicted80 <- mounds[mounds$TRAP_Code%nin%predicted80$TRAP_Code,] # 768 not predicted

# Inspect
length(unique(predicted80$TRAP_Code)) #  unique mounds
length(grid80$id) #   cells with 80%p in the whole Kazanlak
length(survey_grid80) # cells with 80%p in survey area

# How many gridcells with 80% probability actually contain mounds? 
grids_n_mounds8 <- predicted80 %>% 
  group_by(id) %>% 
  count()

length(unique(grids_n_mounds8$id)) #   grids within study area catch mounds

hist(grids_n_mounds8$n, breaks = 40, 
     xlab = "Mound count in grid",
     ylab = "Grid count",
     main = paste0("Number of predicted mounds (n = ",length(unique(predicted80$TRAP_Code)), ") per grid cell ( ",nrow(grids_n_mounds8)," TP) \n among a total of ",length(survey_grid80)," grid cells in the study area with 80% likelihood"))
```

Among 33 grid cells with 80%+ p of containing a mound, 4 were correct(12%), detecting 5 out of 773 mounds (0.64%).

## False Negatives at 80% threshold --- COMPLETION NEEDED
