---
title: "Testing the CNN results after image handpicking in 2022 "
author: "Adela Sobotkova"
date: "November 24, 2021 updated `r format(Sys.time(), '%B %d, %Y')`" 
output: 
  html_document:
    toc: true
    toc_float: true
    highlight: tango
    code_folding: hide
---

## Libraries
```{r libraries}
library(tidyverse)
library(sf)
library(raster)
```

## Load Data

```{r}
objects_needed <- c("kaz",           # aggregate satellite image
                    "cnn_df",        # dataframe of mound predictions for the image 
                    "cnn60_pt",      # point file with predictions 60%+    
                    "grid60",        # spatial grid from 60%+ predictions 
                    "mounds",        # mound points documented in the survey area (773)
                    "survey_ch",     # survey area for verification reduced
                    "survey",        # survey area complete
                    "survey_grid60")# grids with 60+% mound likelihood in the survey area 
script <- c("CNN01_LoadData2022.R")

for (i in objects_needed){
  if (exists(i)){
      print(i)
  }  else {
    source(script)
  #  source(paste0("scripts/",(script[contains(gsub("m","", i), vars = script)])))
  }
}
```
Initial summary: The area of 637 sq km of the aggregated satellite image for the Kazanlak Valley was divided into `dim(cnn_df)[1]`grids of 150 by 150m.  Within these, the CNN model predicted a mound with 60%+ probability in `dim(cnn60_pt)[1]` grids of 150 by 150m (`dim(cnn60_pt)[1]/dim(cnn_df)[1]*100` %). Out of these grids with 60%+ likelihood of mound, `length(survey_grid60)` grids were in the study area of 85 sq km for which we have ground truthed a dataset of 773 mounds, roughly half of them visible in the imagery. 

In the following script, we will use these 430 grid cells and 773 mounds (350 visible) for validation 

## Plot the data
```{r view}
# Plot the 
plot(grid60["mound_probability"])
plot(grid80["mound_probability"])

# View all mounds
plotRGB(kaz, stretch= "lin");
plot(survey_ch, border = "green", lwd = 2, add = TRUE);
#plot(survey_sm, border = "red", lwd = 2, add = TRUE);
#plot(survey$geometry, col = "lightyellow",border = "lightyellow", add = TRUE );
plot(mounds$geometry, add = TRUE, col = "hotpink");
plot(grid80$geometry, add = TRUE, col = "white");
plot(grid60$geometry, add = TRUE, border = "white");
legend("bottom", legend = "Grid cells with 60% (white border) and 80% (white fill) probability of mounds overlayed on \nthe satellite image and survey study area outlines with mounds (pink circles)")

```
The map shows that predictions are distributed randomly through the `dim(kaz)[1]*dim(kaz)[2]*0.000001` sq km area of the survey zone (NE of the actual image) without the clustering that characterizes the mound dataset. The selection by the CNN is a bit indiscriminate of land use, which is especially surprisingly within the homogeneous water surface of the Koprinka reservoir (how is this happening!) and undercounting in the area of the Yasenovo necropolis in the NW of the study area. The 2021 numbers seemed more accurate in that respect?


## Validation

The Tundzha Regional Archaeological project surveyed 85 sq km of the area included in the satellite imagery in 2009-2011, documenting the burial mounds within which create a dataset suitable for validation.

Area surveyed contains 773 documented mounds of various shapes and sizes, circa half of them visible in the satellite imagery (precise number is difficult to establish as the visibility gets fuzzy). This area is covered by 430 of the 897 grid cells which have 0.6+ probability of containing a mound.
Let's explore how many mounds are found inside these 0.6+p grid cells and which mounds (small, large, .etc.) these are. Bring in mound attributes and check correlation. Does size play a role?



## True Positives at 60% threshold
Archaeological survey documented 773 mounds in the study area, half of which are visible by naked eye on the surface. Mounds from this dataset that fall in the 60%+ probability grid cells form the true positives and represent the recall(?) of the CNN model.

```{r overlap}
# 22 out of 773 mounds are found in 60%+ grid cells
predictedat60 <- st_intersection(mounds, grid60)


# Check in which gridcells the predicted mounds are located
success60 <- grid60 %>% 
  slice(which(lengths(st_contains(grid60, mounds))>0)) # %>% distinct()
  #slice(which(lengths(st_intersects(grid60, mounds))>0))
length(unique(success60$id))
plot(success60)

# Which mounds are not-predicted? 
`%nin%` = Negate(`%in%`)
unpredicted60 <- mounds[mounds$TRAP_Code%nin%predictedat60$TRAP_Code,] # 706 not predicted
```

`dim(predictedat60)[1]` mounds of the 773 intersect `length(unique(success60$id))` of the 430 grids with 60+% probability of containing a mound. `dim(unpredicted60)[1]` mounds across remain unpredicted. With 773 burial mounds and at least 350 visible ones, this produces a  `dim(predictedat60)[1]/773*100` % success rate on total and `dim(predictedat60)[1]/350*100` % on visible features.

Having checked the mounds dataset, let's now check how many of the 60%+ grids were successful in containing mounds. 
And afterwards, are smaller or bigger mounds being predicted? 

```{r tp-at-60}
# Inspect
length(unique(predictedat60$TRAP_Code)) # 22 unique mounds

# Summarize how many mounds each TP gridcell contains 
grids_n_mounds <- predictedat60 %>% 
  group_by(id) %>% 
  count()

length(unique(grids_n_mounds$id)) # 16 unique grids within study area catch the 22 mounds; consistent with success60 above

hist(grids_n_mounds$n, 
     #breaks = 40, 
     xlab = "Mound count in grid",
     ylab = "Grid count",
     main = paste0("Distribution of mounds (n = ",length(unique(predictedat60$TRAP_Code)), ") across ",nrow(grids_n_mounds)," of the ",length(survey_grid60),"  \n grid cells with 60+% probability of mound within survey area"))

```
Summary: `nrow(grids_n_mounds)` of the `length(survey_grid60)` grid cells with 60+% probability of mound within survey area contain between 1 and 4 known mounds.
The multiples make sense (and are on the conservative side) for the necropolis of little 10m diameter mounds in the NW

## Precision
In the 290 cells with 60%+ likelihood of mound, there were 16 successful cells (`16/290*100` %) and 274 false positive cells (`(290-16)/290*100`). Out of the 773 known mounds and 350 visible mounds, 22 fell in the successful cells (`22/350*100`% success rate) and `350-22` were undetected (`328/350*100` %).

The numbers are worse than 2021 with the larger dataset. I wonder what is being detected? Are we using the NIR band? (Which the human readers were using as a verification)

Continue:::
The previous chunk shows that 22 of 773 mounds appear in 16 of the 290 grids with 60%+ probability of mounds and intersect with the TRAP study area. 




### Write prediction status to mounds
```{r write-predicted60-status}
# Add the information on prediction to mound dataset
mounds <- mounds %>% 
  mutate(predictedat60 =  case_when(TRAP_Code %in% predictedat60$TRAP_Code ~ "yes",
                                  TRAP_Code %nin% predictedat60$TRAP_Code ~ "no"
                                  ))
```



## True Positives at 80% threshold  --- COMPLETION NEEDED
## False Negatives at 60% threshold
Given 67 mounds were predicted by CNN, where are the remaining 627?
Let's explore where the 627 missing mounds are and what probability has been assigned to the cells that contain them by CNN model.

```{r recreate-complete-grid}
side <- 150
y <- cnn_df$coord_y
x <- cnn_df$coord_x


# define the plot edges based upon the plot radius. 
yPlus <- y+side
xPlus <- x+side

# calculate polygon coordinates for each plot centroid. 
square <- cbind(x,yPlus,  # NW corner
                xPlus, yPlus,  # NE corner
                xPlus,y,  # SE corner
                x,y, # SW corner
                x,yPlus)  # NW corner again - close polygon

# Extract the image ID information
ID <- cnn_df$`Image filename`

# create spatial polygons (squares) from mound coordinates with mapply
polys <- SpatialPolygons(mapply(function(poly, id) 
{
  xy <- matrix(poly, ncol=2, byrow=TRUE)
  Polygons(list(Polygon(xy)), ID=id)
}, 
split(square, row(square)), ID),
proj4string=CRS(as.character("+proj=utm +zone=35 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")))

# Convert to sf feature via a SpatialDataframe (to preserve mound IDs)
polys_df <- SpatialPolygonsDataFrame(polys, data.frame(id=ID, row.names = ID,
                                                       mound_probability = 1 - cnn_df$`Raw Prediction`))
polys_df
cnnall_grid <- st_as_sf(polys_df)

# Write to shapefile
st_write(cnnall_grid, "output_data/cnngrid_150.shp",driver = 'ESRI Shapefile', append=FALSE)

# Remove interim files
remove(polys,polys_df, square, x,y,xPlus,yPlus,side)

```

```{r explore-FN}
# How many mounds are undetected and 
# Get info on regarding prediction from the overall grid cnn_all
missing60 <- st_intersection(unpredicted60, cnnall_grid) # 744 mounds undetected

# Crop master grid by survey area 
survey_grid <- st_intersection(survey_ch, cnnall_grid)

# Which grids are they in (with what probablity?)
FN <- cnnall_grid %>%   # 383 grids are False Negatives
  slice(which(lengths(st_contains(cnnall_grid, unpredicted60))>0)) 

# Probability overview
hist(FN$mound_probability)

# Summarize how many mounds are there per each FN gridcell 
grids_FN_mounds <- missing60 %>% 
  group_by(id) %>% 
  count()

length(unique(grids_FN_mounds$id)) # 383 FN grids

hist(grids_FN_mounds$n, 
     #breaks = 40, 
     xlab = "Mound count per FN gridcell",
     ylab = "Grid count",
     main = paste0("False negatives: distribution of mounds (n = ",length(unique(missing60$TRAP_Code)), ") across ",nrow(grids_FN_mounds)," of the ",length(survey_grid),"  \n grid cells within survey area"))
```

Undetected mounds appear in 383 gridcells, which contain between 1 and 20 mounds, heavily skewed towards small mound number within cell. 

What is the probability of the cells that contain undetected mounds?
```{r missing-prob}
# What is the mound_probability in these gridcells?
hist(missing60$mound_probability, main = "Probability of gridcells containing undetected mounds")
```
## False Negatives at 80% threshold --- COMPLETION NEEDED
## Visualize results 
```{r mapview}
# See the predicted mounds over grids with 60% likelihood of mound plus unpredicted mounds

library(mapview)
mapview(unpredicted60, color = "orange", alpha= 0.5,
         map.types = c("Esri.WorldImagery", "CartoDB.Positron"),
         layer.name = "Undetected mounds")+
  mapview(grids_n_mounds, zcol = "n", at = c(1, 2, 5, 10, 15, 30),
          layer.name = "Detected mound no per gridcell")+
  mapview(survey_grid60, layer.name = "Gridcells with 60%+ probability")
```

### Explore impact of height on detectability
Let's look at whether mound Height is a factor impacting detectability. (it is a factor for visual inspection as higher mounds are more visible in sat img.) 
```{r explore-height}
# filter the mounds for largesh and noticeable ones (2+ meters)
large <- mounds %>% 
  filter(Height>=2) # 247 obs

mounds %>% 
  group_by(predicted60) %>% 
  summarize(min = min(Height), max = max(Height))

# Check the height difference between un- and predicted mounds
boxplot(Height~predicted60, data = mounds, xlab = "Was this field-confirmed mound detected with CNN?")

# Check height distribution btw un- and predicted mounds
hist(mounds$Height[mounds$predicted60 == "no"], col = "pink", breaks  = 20,main = "Histogram of detected (yellow) and undetected (pink) mounds by height",
     xlab = "Mound height (m)");
hist(mounds$Height[mounds$predicted60 == "yes"], col = "yellow", add =TRUE, alpha = 0.5);
```

### T-test on height
```{r ttest-height}
# Is there a significant relationship between the Height of a mound and its detection?
t.test(Height ~ as.factor(predicted60), data = mounds)
str(t.test(Height ~ as.factor(predicted60), data = mounds))
# ... the answer seems to be yes with p at 3.14e-05
```

## Summary of success rates - 2021 results

146 mounds was detected out of 773, making the CNN predict (over 60% probability) ca 20% of mounds documented in the study area. Out of the 192 gridcells which are within the archaeological study area and have 60%+ probabilty of containing a mound, 50 actually contained between 1 and 29 mounds. 
Smaller mounds seem to enjoy higher detection rate according to the t-test, but that may not be the right metric. Let's say that surprisingly, the large mounds are not detected as frequently as I would expect, and as human would detect them.

The remaining undetected 620 mounds were found in 259 gridcells, clustered from 1 to 28 specimen. Their exploration shows a problem with underlying grid overlap which cause reduplication of probability values (which are not identical) and proliferation of mound duplicates which each have different probability value! See section `discrepancy` above.

Once overlap is resolved, I can slide the probability downwards and downwards and see if I get higher veracity (more cells will be TPs) with higher probability or decrease missing values (fewer FNs) with lower probability cells.

### What is the model succeeding at and what not?
Look at the NW cluster and see that many mounds sit on the border of polygons, and so get counted twice > ca 20 mounds, so the numbers of predicted are slightly exaggerated
In NE we are missing a lot of very large 20m diameter+ mounds that are covered with scrub (look forested) 
In the south a lot of forest patches and beach boundaries are getting picked up that are false positives. Here the overlap may be at play !! REVISIT.
All in all the model seems to pick on the bright signatures rather than round shapes and there is an odd combination of large forested lozenges and small bare spots picked up, but somehow round forested images with little excavated (bright) trenches on the south side is not getting picked up.



## Additional visualisations

```{r viz-mounds-cnn}

# Where are the mounds vis-a-vis the probability values
cnnall_grid %>% 
  ggplot()+
  geom_sf(aes(color = mound_probability))+
  scale_color_gradient(low = "green", 
                       high = "red",
                       "Mound probability")+
  geom_sf(data = mounds) +
  ggtitle("Location of known mounds vis-a-vis the CNN predictions")


# See the 146/166 mound features predicted with 60%+ probability 
grid60 %>% 
 # filter(mound_probability>0.5) %>% 
  ggplot()+
  geom_sf(aes(color = mound_probability))+
  scale_color_gradient(low = "green", 
                       high = "red",
                       "Mound probability")+
  geom_sf(data = mounds$geometry[mounds$predicted60 == "yes"], size = mounds$Height[mounds$predicted60 == "yes"], col = "lightgrey", alpha = 0.5) +
  ggtitle("Predicted mounds in Kazanlak")

# See the 627 mounds not caught by the cells with 60%+ probability 
grid60 %>% 
  # filter(mound_probability>0.5) %>% 
  ggplot()+
  geom_sf(aes(color = mound_probability))+
  scale_color_gradient(low = "green", 
                       high = "red",
                       "Mound probability")+
  #geom_sf(data = unpredicted60$geometry)+
  geom_sf(data = mounds$geometry[mounds$predicted60 == "no"], size = mounds$Height[mounds$predicted60 == "no"], col = "lightgrey", alpha = 0.5)+
  ggtitle("Unpredicted mounds in Kazanlak")
```

## Side-by-side views 
These images take mound height into account and split the results into facets
```{r ggplot-facet}
# Cumulative facetted visual

ggplot()+
  geom_sf(data = grid60, aes(fill = mound_probability)) +
  labs(fill = "Probability of mound")+
  # add mounds
  geom_sf(data = mounds, 
          aes(size = Height, col = "red",  alpha = 0.6)) +
  facet_wrap(~predicted60)+
  theme_bw() + 
  guides(color = FALSE,
         size = guide_legend(order = 1),
         fill = guide_legend(order = 2),
         alpha = FALSE)+
  ggtitle("Mounds located in gridcells with 60%+ probability (no=620, yes=146)")
```