---
title: "Testing the CNN results after image handpicking in 2022 "
author: "Adela Sobotkova"
date: "November 24, 2021 updated `r format(Sys.time(), '%B %d, %Y')`" 
output: 
  html_document:
    toc: true
    toc_float: true
    highlight: tango
    code_folding: hide
---

## Libraries
```{r libraries}
library(tidyverse)
library(sf)
library(raster)
```

## Load Data

```{r}
objects_needed <- c("kaz",           # aggregate satellite image
                    "cnn_df",        # dataframe of mound predictions for the image 
                    "cnn60_pt",      # point file with predictions 60%+    
                    "grid60",        # spatial grid from 60%+ predictions 
                    "mounds",        # mound points documented in the survey area (773)
                    "survey_ch",     # survey area for verification reduced
                    "survey",        # survey area complete
                    "survey_grid60")# grids with 60+% mound likelihood in the survey area 
script <- c("CNN01_LoadData2022.R")

for (i in objects_needed){
  if (exists(i)){
      print(i)
  }  else {
    source(script)
  #  source(paste0("scripts/",(script[contains(gsub("m","", i), vars = script)])))
  }
}
```
Initial summary: The area of 637 sq km of the aggregated satellite image for the Kazanlak Valley was divided into `dim(cnn_df)[1]`grids of 150 by 150m.  Within these, the CNN model predicted a mound with 60%+ probability in `dim(cnn60_pt)[1]` grids of 150 by 150m (`dim(cnn60_pt)[1]/dim(cnn_df)[1]*100` %). Out of these grids with 60%+ likelihood of mound, `length(survey_grid60)` grids were in the study area of 85 sq km for which we have ground truthed a dataset of 773 mounds, roughly half of them visible in the imagery. 

In the following script, we will use these 430 grid cells and 773 mounds (350 visible) for validation 

## Plot the data
```{r view}
# Plot the 
plot(grid60["mound_probability"])
plot(grid80["mound_probability"])

# View all mounds
plotRGB(kaz, stretch= "lin");
plot(survey_ch, border = "green", lwd = 2, add = TRUE);
#plot(survey_sm, border = "red", lwd = 2, add = TRUE);
#plot(survey$geometry, col = "lightyellow",border = "lightyellow", add = TRUE );
plot(mounds$geometry, add = TRUE, col = "hotpink");
plot(grid80$geometry, add = TRUE, col = "white");
plot(grid60$geometry, add = TRUE, border = "white");
legend("bottom", legend = "Grid cells with 60% (white border) and 80% (white fill) probability of mounds overlayed on \nthe satellite image and survey study area outlines with mounds (pink circles)")

```
The map shows that predictions are distributed randomly through the `dim(kaz)[1]*dim(kaz)[2]*0.000001` sq km area of the survey zone (NE of the actual image) without the clustering that characterizes the mound dataset. The selection by the CNN is a bit indiscriminate of land use, which is especially surprisingly within the homogeneous water surface of the Koprinka reservoir (how is this happening!) and undercounting in the area of the Yasenovo necropolis in the NW of the study area. The 2021 numbers seemed more accurate in that respect?


## Validation

The Tundzha Regional Archaeological project surveyed 85 sq km of the area included in the satellite imagery in 2009-2011, documenting the burial mounds within which create a dataset suitable for validation.

Area surveyed contains 773 documented mounds of various shapes and sizes, circa half of them visible in the satellite imagery (precise number is difficult to establish as the visibility gets fuzzy). This area is covered by 430 of the 897 grid cells which have 0.6+ probability of containing a mound.
Let's explore how many mounds are found inside these 0.6+p grid cells and which mounds (small, large, .etc.) these are. Bring in mound attributes and check correlation. Does size play a role?



## True Positives at 60% threshold
Archaeological survey documented 773 mounds in the study area, half of which are visible by naked eye on the surface. Mounds from this dataset that fall in the 60%+ probability grid cells form the true positives and represent the recall(?) of the CNN model.

```{r overlap}
# 22 out of 773 mounds are found in 60%+ grid cells
predictedat60 <- st_intersection(mounds, grid60)


# Check in which gridcells the predicted mounds are located
success60 <- grid60 %>% 
  slice(which(lengths(st_contains(grid60, mounds))>0)) # %>% distinct()
  #slice(which(lengths(st_intersects(grid60, mounds))>0))
length(unique(success60$id))
plot(success60$geometry, main = "60%+ p grids that actually contain mounds")

# Which mounds are not-predicted? 
`%nin%` = Negate(`%in%`)
unpredicted60 <- mounds[mounds$TRAP_Code%nin%predictedat60$TRAP_Code,] # 706 not predicted
```

`dim(predictedat60)[1]` mounds of the 773 intersect `length(unique(success60$id))` of the 430 grids with 60+% probability of containing a mound. `dim(unpredicted60)[1]` mounds across remain unpredicted. With 773 burial mounds and at least 350 visible ones, this produces a  `dim(predictedat60)[1]/773*100` % success rate on total and `dim(predictedat60)[1]/350*100` % on visible features.

Having checked the mounds dataset, let's now check how many of the 60%+ grids were successful in containing mounds. 
And afterwards, are smaller or bigger mounds being predicted? 

```{r tp-at-60}
# Inspect
length(unique(predictedat60$TRAP_Code)) # 22 unique mounds

# Summarize how many mounds each TP gridcell contains 
grids_n_mounds <- predictedat60 %>% 
  group_by(id) %>% 
  count()

length(unique(grids_n_mounds$id)) # 16 unique grids within study area catch the 22 mounds; consistent with success60 above

hist(grids_n_mounds$n, 
     #breaks = 40, 
     xlab = "Mound count in grid",
     ylab = "Grid count",
     main = paste0("Distribution of mounds (n = ",length(unique(predictedat60$TRAP_Code)), ") across ",nrow(grids_n_mounds)," of the ",length(survey_grid60),"  \n grid cells with 60+% probability of mound within survey area"))

```
Summary: `nrow(grids_n_mounds)` of the `length(survey_grid60)` grid cells with 60+% probability of mound within survey area contain between 1 and 4 known mounds.
The multiples make sense (and are on the conservative side) for the necropolis of little 10m diameter mounds in the NW

## Precision
In the 290 cells with 60%+ likelihood of mound, there were 16 successful cells (`16/290*100` %) and 274 false positive cells (`(290-16)/290*100`). Out of the 773 known mounds and 350 visible mounds, 22 fell in the successful cells (`22/350*100`% success rate) and `350-22` were undetected (`328/350*100` %).

The numbers are worse than 2021 with the larger dataset. I wonder what is being detected? Are we using the NIR band? (Which the human readers were using as a verification)

Continue:::
The previous chunk shows that 21 of 773 mounds appear in 15 of the 288 grids with 60%+ probability of mounds and intersect with the TRAP study area. 

## False Negatives at 60% threshold
Given 21 mounds were predicted by CNN, where are the remaining ~750?
Let's explore where the missing mounds are and what probability has been assigned to the cells that contain them by CNN model.

```{r recreate-complete-grid}
side <- 150
y <- cnn_df$coord_y
x <- cnn_df$coord_x


# define the plot edges based upon the plot radius. 
yPlus <- y+side
xPlus <- x+side

# calculate polygon coordinates for each plot centroid. 
square <- cbind(x,yPlus,  # NW corner
                xPlus, yPlus,  # NE corner
                xPlus,y,  # SE corner
                x,y, # SW corner
                x,yPlus)  # NW corner again - close polygon

# Extract the image ID information
ID <- cnn_df$`Image filename`

# create spatial polygons (squares) from mound coordinates with mapply
polys <- SpatialPolygons(mapply(function(poly, id) 
{
  xy <- matrix(poly, ncol=2, byrow=TRUE)
  Polygons(list(Polygon(xy)), ID=id)
}, 
split(square, row(square)), ID),
proj4string=CRS(as.character("+proj=utm +zone=35 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")))

# Convert to sf feature via a SpatialDataframe (to preserve mound IDs)
polys_df <- SpatialPolygonsDataFrame(polys, data.frame(id=ID, row.names = ID,
                                                       mound_probability = 1 - cnn_df$`Raw Prediction`))
polys_df
cnnall_grid <- st_as_sf(polys_df)

# Write to shapefile
#st_write(cnnall_grid, "output_data/cnngrid_150.shp",driver = 'ESRI Shapefile', append=FALSE)

# Remove interim files
remove(polys,polys_df, square, x,y,xPlus,yPlus,side)

```

```{r explore-FN}
# How many mounds are undetected and 
# Get info on regarding prediction from the overall grid cnn_all
missing60 <- st_intersection(unpredicted60, cnnall_grid) # 744 mounds undetected

# Crop master grid by survey area 
survey_grid <- st_intersection(survey_ch, cnnall_grid)

# Which grids are they in (with what probablity?)
FN <- cnnall_grid %>%   # 383 grids are False Negatives
  slice(which(lengths(st_contains(cnnall_grid, unpredicted60))>0)) 

# Probability overview
hist(FN$mound_probability)

# Summarize how many mounds are there per each FN gridcell 
grids_FN_mounds <- missing60 %>% 
  group_by(id) %>% 
  count()

length(unique(grids_FN_mounds$id)) # 383 FN grids

# Histograms of the FN grids and their probability
hist(grids_FN_mounds$n, 
     #breaks = 40, 
     xlab = "Mound count per FN gridcell",
     ylab = "Grid count",
     main = paste0("False negatives: distribution of mounds (n = ",length(unique(missing60$TRAP_Code)), ") across ",nrow(grids_FN_mounds)," of the ",length(survey_grid),"  \n grid cells within survey area"))
```

Undetected mounds appear in 383 gridcells, which contain between 1 and 20 mounds, heavily skewed towards small mound number within cell. 

What is the probability of the cells that contain undetected mounds?
```{r missing-prob}
# Histograms of the probability associated with the undetected mounds (same thing as above, only higher no)
hist(missing60$mound_probability, main = "Probability of gridcells containing undetected mounds")
```


### Write prediction status to mounds
```{r write-predicted60-status, eval = FALSE}
# Add the information on prediction to mound dataset
mounds <- mounds %>% 
  mutate(predictedat60 =  case_when(TRAP_Code %in% predictedat60$TRAP_Code ~ "yes",
                                  TRAP_Code %nin% predictedat60$TRAP_Code ~ "no"
                                  ))
```

## Visualize results 
```{r mapview}
# See the predicted mounds over grids with 60% likelihood of mound plus unpredicted mounds

library(mapview)
mapview(unpredicted60, color = "orange", alpha= 0.5,
         map.types = c("Esri.WorldImagery", "CartoDB.Positron"),
         layer.name = "Undetected mounds")+
  mapview(grids_n_mounds, zcol = "n", at = c(1, 2, 5, 10, 15, 30),
          layer.name = "Detected mound no per gridcell")+
  mapview(survey_grid60, layer.name = "Gridcells with 60%+ probability")
```


## True Positives at 80% threshold  --- COMPLETION NEEDED

## False Negatives at 80% threshold --- COMPLETION NEEDED


### Explore impact of height on detectability
Let's look at whether mound Height is a factor impacting detectability. (it is a factor for visual inspection as higher mounds are more visible in sat img.) 
```{r explore-height}
# filter the mounds for largesh and noticeable ones (2+ meters)
large <- mounds %>% 
  filter(Height>=2) # 247 obs

mounds %>% 
  group_by(predictedat60) %>% tally() 
  summarize(min = min(Height), max = max(Height))


# Check the height difference between un- and predicted mounds
boxplot(Height~predictedat60, data = mounds, xlab = "Was this mound detected with CNN?")

# Check height distribution btw un- and predicted mounds
hist(mounds$Height[mounds$predictedat60 == "no"], col = "pink", breaks  = 20,main = "Histogram of detected (yellow) and undetected (pink) mounds by height",
     xlab = "Mound height (m)");
hist(mounds$Height[mounds$predictedat60 == "yes"], col = "yellow", add =TRUE, alpha = 0.5);
```

### T-test on height
```{r ttest-height}
# Is there a significant relationship between the Height of a mound and its detection?
t.test(Height ~ as.factor(predictedat60), data = mounds)
str(t.test(Height ~ as.factor(predictedat60), data = mounds))
# ...  p at 0.2497 does not inspire confidence
```

## Summary of success rates - 2021 results

21 mounds was detected out of 773, 752 remain undetected. Out of the 288 gridcells which are within the archaeological study area and have 60%+ probability of containing a mound, 15 actually contained between 1 and 4 mounds. 

The remaining undetected 752 mounds were found in 384 gridcells with probability below 60%, clustered from 1 to 28 specimen. 


### What is the model succeeding at and what not?
NW cluster is not getting detected unlike in 2021 dataset.
Large mounds under and around Shipka are not getting detected.
Grids with 60%+p appear in the middle of the reservoir, which makes the 'what is the CNN detecting' a big question

All in all, I cannot see a logic behind the model (unlike in 2021, when the model seemed to pick on the bright signatures of roads). The round shapes don't seem to be it.

## Side-by-side views 
These images take mound height into account and split the results into facets
```{r ggplot-facet}
# Cumulative facetted visual

ggplot()+
  geom_sf(data = grid60, aes(fill = mound_probability)) +
  labs(fill = "Probability of mound")+
  # add mounds
  geom_sf(data = mounds, 
          aes(size = Height, col = "red",  alpha = 0.6)) +
  facet_wrap(~predictedat60)+
  theme_bw() + 
  guides(color = FALSE,
         size = guide_legend(order = 1),
         fill = guide_legend(order = 2),
         alpha = FALSE)+
  ggtitle(paste0("Mounds located in gridcells with 60%+ 
          probability (no=",nrow(predictedat60),",yes=",nrow(missing60),")"))
```