---
title: "Cut out Images for CNN"
author: "Adela Sobotkova"
date: "11/18/2021 updated `r format(Sys.time(), '%B %d, %Y')`" 
output: 
  html_document:
    toc: true
    toc_float: true
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and data
```{r libraries}
library(sf)
library(raster)
source("STAMPS_LoadLargeImg.R")
```

## Need better training data?


Let's filter the mounds to the large and visible ones, and prepare cutouts so they can be run through the model and see if the detection rate improves.
First I create the loop to produce squares/grids to use to cutout segments from the raster, then I need to figure out by what characteristics I best improve the model (e.g. get those large scrubby mounds from NE)
```{r mound-grids-polys}

radius <- 75 #twice that is 150m per side of polygon
radius <- 125 #twice that is 250m per side of polygon
mounds$northing <- st_coordinates(mounds)[,2]
mounds$easting <- st_coordinates(mounds)[,1]


# define the plot edges based upon the plot radius. 

yPlus <- mounds$northing+radius
xPlus <- mounds$easting+radius
yMinus <- mounds$northing-radius
xMinus <- mounds$easting-radius

# calculate polygon coordinates for each plot centroid. 
square <- cbind(xMinus,yPlus,  # NW corner
	xPlus, yPlus,  # NE corner
	xPlus,yMinus,  # SE corner
	xMinus,yMinus, # SW corner
	xMinus,yPlus)  # NW corner again - close polygon

# Extract the mound ID information
ID <- mounds$TRAP_Code

# create spatial polygons (squares) from mound coordinates with mapply
polys <- SpatialPolygons(mapply(function(poly, id) 
		{
	  xy <- matrix(poly, ncol=2, byrow=TRUE)
	  Polygons(list(Polygon(xy)), ID=id)
	  }, 
	split(square, row(square)), ID),
	proj4string=CRS(as.character("+proj=utm +zone=35 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")))

plot(polys)

# Convert to sf feature via a SpatialDataframe (to preserve mound IDs)
polys_df <- SpatialPolygonsDataFrame(polys, data.frame(id=ID, row.names=ID))
polys_df
mnd_poly <- st_as_sf(polys_df)
mnd_poly

# Add attributes
mnd_poly <- mnd_poly %>% 
  left_join(mounddata, by = c("id"="MoundID"))

# Write to shapefile
st_write(mnd_poly, "output_data/moundpolys250.shp",driver = 'ESRI Shapefile', append=FALSE)

# Remove interim files
remove(polys,polys_df, square)

# Look at the polygons
mnd_poly2m <- mnd_poly %>% 
  filter(Height>=2) 
plot(mnd_poly$geometry, main = "All mounds (black) and 2m+ mounds (red)"); 
plot(mnd_poly2m$geometry, col = "red", add = TRUE)
```
### Find Features Visible on the Ground (in IKONOS imagery)
To obviate having to look at all mounds in sat img, I will use the RS Analysis geodatabase that students filled out when doing visual remote sensing. Each 'anomaly' has F_Type, which the analyst uses to specify what s/he is seeing in the sat image. The categories are: 

- VIS: Visible on surface
- CCM: Combination crop mark 
- CSH: Crop and shadow mark
- PCM: Positive crop or weed mark
- NCM: Negative crop or weed mark
- SM: soil mark
- SH: shadow mark
- UN: unknown or other

Picking up 'Visible on Surface' features, that spatially overlap with registered mounds should bring up the best possible training dataset. (especially if we also cross with mounds higher than 2 m)

```{r find-visible}
# Read in all remotely-sensed features 
rs <- st_read("~/Desktop/TRAP_Oxbow/KAZ/RS_Points12.shp")

# Filter for the confirmed mounds and see breakdown of Type
rs %>% 
  filter(F_ACaus == "MND") %>% # we filter for actual mounds
  group_by(F_Type) %>% 
  count()  # code VIS denotes 152 "visible on surface" mounds 

# Look at the confirmed mounds labelled as 'visible on surface'
library(mapview)
rs %>% 
  filter(F_ACaus == "MND" & F_Type == "VIS") %>% 
  mapview()

```
Now that we see how the RS_point dataset encodes mounds visible on surface in the IKONOS image, can we create a separate dataset out of the visible mounds and use it (especially the TRAP_code column) to filter the mounds dataset. We shall pick fairly high mounds (3.5m) to have a reliable dataset.

```{r filter-visible-high}
visible_mounds <- rs %>% 
  filter(F_ACaus == "MND" & F_Type == "VIS") 

mounds %>% 
  filter(Height>=3) %>%  # 185 are 2+ m high, 179 are 3+, 147 are 3.5+, 143 are 4+m 
  filter(TRAP_Code %in% visible_mounds$TRAP_code) %>%  #72 2.5 features, or 68 3m features, or 56 3.5 features caught 
  mapview()

visiblemounds3_5 <- mounds %>% 
  # 185 are 2+ m high, 179 are 3+, 147 are 3.5+, 143 are 4+m 
  filter(Height>=3.5) %>% 
  # 72 2.5 features, or 68 3m features, or 56 3.5 features caught 
  filter(TRAP_Code %in% visible_mounds$TRAP_code)

visiblemounds3_5$TRAP_Code
```


### Check that visible mounds are within the Kaz raster
If we extract values from raster at mound point and the result is NA, than the mound does not overlap with the raster. Check for NAs and eliminate TRUE values. 
```{r}
is.na(extract(KAZcropped[[1]],visiblemounds3_5))
```
All mounds are within the satellite image. 

### Loop Proof of Concept: : Small stamps from low-resolution satellite image

```{r loop-test, eval=FALSE}
# Loop over a sample of largish mounds and cutout postage stamp rasters
randomID <- mnd_poly2m$id%>% 
  sample(size = 25, replace = FALSE)
unique(randomID)
# Better: Loop over visible 3.5m high mound IDs 
randomID <- visiblemounds3_5$TRAP_Code

# Create a directory for stamps
dir.create("stamps_low_res")

# Create cutouts per sample number
for (i in randomID){
  new_clip <- crop(kaz, mnd_poly[mnd_poly$id ==i,])
  plotRGB(new_clip)
  writeRaster(new_clip, filename= paste0("stamps_low_res/",i,".tiff"), format = "GTiff", overwrite =TRUE)
}

```

The clipping works only if there is overlap with the raster! Non-overlapping polygons don't work so well, asyou can see if you randomly choose an ID. Such as  `mnd_poly[mnd_poly$id ==2006,]`



### More Defensive Loop

The process of generating stamps from raster with polygon outlines consists of two steps:
1) check if a given polygon overlaps fully with the raster
2) crop raster if it does, skip to next if it doesn't

To check spatial overlap, gContainsProperly() function from geos library is needed as intersection in sf does not work here. I therefore need to convert to 'Spatial' format for the validation.
See more at https://gis.stackexchange.com/questions/34535/detect-whether-there-is-a-spatial-polygon-in-a-spatial-extent

```{r generate--slowly-stamps-that-overlap-raster, eval = FALSE}
# Check spatial overlap - 
# in <- intersect(e, extent(sp))
# if (isTRUE(in)) { in <- gIntersects(as(e, 'SpatialPolygons'), sp) }

# Validation and cropping in a loop
library(rgeos)
for (i in randomID){
  lapply(poly, function(raster) {
    poly <- as(mnd_poly[mnd_poly$id ==i,], "Spatial")
    ei <- as(extent(kaz), "SpatialPolygons")
    if (gContainsProperly(ei,poly)) {
      paste0(i," is fully within")
      new_clip <- crop(kaz, mnd_poly[mnd_poly$id ==i,])
      plotRGB(new_clip)
      writeRaster(new_clip, 
                  filename= paste0("stamps_low_res/",i,".tiff"), 
                  format = "GTiff", overwrite =TRUE)
  } else if (gIntersects(ei,poly)) {
    paste0(i," intersects")
  } else {
    paste0(i," is fully outside the raster")
  }
})
}

# Review results
i
plotRGB(new_clip, stretch = "lin")

# This works, but stamps are pixellated. Higher resolution is needed
```

### Final Loop: using hi-res original merged image
The satellite image suffers from histogram imbalance, however, that should aid with augmentation, so I won't worry about it now.

```{r large-stamps}
# If large raster is not in memory
# setwd("E:/TRAP Workstation/Shared GIS/Satellite imagery/IKONOS/Kazanlak/ERDAS/Cormac")
# KAZcropped <- brick("Kazcropped.tif")
# Alternatively if a rebuild is needed, run
#source("MOSAIC.R")

# Loop to check for polygon and raster overlap and to generate stamps 
for (i in randomID){
  lapply(poly, function(raster) {
    poly <- as(mnd_poly[mnd_poly$id ==i,], "Spatial")    # convert polygon to 'Spatial' format
    ei <- as(extent(KAZcropped), "SpatialPolygons")             # create an extent of sat img as 'Spatial' format
    if (gContainsProperly(ei,poly)) {
      paste0("Polygon for ", i," is fully within raster")
      new_clip <- crop(KAZcropped, mnd_poly[mnd_poly$id ==i,])  # create the cutout
      #plotRGB(new_clip)
      writeRaster(new_clip,
                  filename= paste0("stamps/",i,".tif"),
                  # filename= paste0("E:/TRAP Workstation/Shared GIS/Satellite imagery/IKONOS/Kazanlak/ERDAS/Cormac/stamps/",i,".tif"),
                  format = "GTiff", overwrite =TRUE)
  } else if (gIntersects(ei,poly)) {
    paste0(i," intersects")
  } else {
    paste0(i," is fully outside the raster")
  }
})
}

plotRGB(new_clip)
```

